---
# title: "Predicting Bird groups based on their bone measurements"
author: "Jouke Profijt"
date: "September 11, 2018"
output: pdf_document

---
```{r}
#Copyright (c) 2018 Jouke Profijt.
#Licensed under GPLv3. See gpl.md
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

```{r}
BirdBones <- read.csv("../data/bird.csv",header = T, sep = ",")
```

# Introduction
# Research Question
   What bone or group of bones that most birds have in common, is the most significant for the function in the diffrent ecological groups?
    
## Data
    Data recieved from  [Birds' Bones and Living Habits](https://www.kaggle.com/zhangjuefei/birds-bones-and-living-habits)
    
    Bone measurements were measured from a skeleton collection of Natural History Museum of Los Angeles County, 
    provided by Dr. D. Liu of beijing Museaum of Natural History
    
# Exploratory Data Analyses





```{r Swimming Birds}
boxplot(SW[2:11], las = 2)

```
I'm creating a graph which displays the bonelengths on y axis an the Id on x colorcoded by their ecological group. by evaluating this we can see if some groups have overall larger or smaller bones and we see if there are big outliers.

```{r}
library(ggplot2)
library(reshape)
source("../scripts/BoneMeans.R")
BirdBones <- BoneMeans(BirdBones)
ggplot(data = BirdBones, aes(id, means, colour = type)) +
  ggtitle("Bone lenghts per Ecological group")+
  ylab("Mean lenght of total bird bones in mm") +
  xlab("Bird Sample")+
  geom_point()
```
As seen above swimming birds have the biggest bones, but also shown is that there are a lot more samples in that group where there is a lot of variation. I can look into cleaning up the data and removing the biggest outliers in this group. Singing birds also have a lot of samples but there is much less variation and so more certanty. 

For the rest of the birds there are not a lot of sample so maby we could try and normalizing the data so there is an even amount of samples per group.

