---
# title: "Predicting Bird groups based on their bone measurements"
author: "Jouke Profijt"
date: "September 11, 2018"
output: pdf_document

---
```{r}
#Copyright (c) 2018 Jouke Profijt.
#Licensed under GPLv3. See gpl.md
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

```{r}
BirdBones <- read.csv("../data/bird.csv",header = T, sep = ",")
```

# Introduction

## Research Question
    Are the bones of birds significantly distinct per group/functioning of birds, and can we use them to identify the ecological group of unknown fossils?
    
## Data
    Data recieved from  [Birds' Bones and Living Habits](https://www.kaggle.com/zhangjuefei/birds-bones-and-living-habits)
    
    Bone measurements were measured from a skeleton collection of Natural History Museum of Los Angeles County, 
    provided by Dr. D. Liu of beijing Museaum of Natural History
    
# Exploratory Data Analyses


```{r Grouping all birds}
# Swimming
SW <- BirdBones[BirdBones$type == "SW",]
# Wading
W <- BirdBones[BirdBones$type == "W",]
# Terrestial
TER <- BirdBones[BirdBones$type == "T",]
# Raptors
R <- BirdBones[BirdBones$type == "R",]
# Scansorial
P <- BirdBones[BirdBones$type == "P",]
# Singing
SO <- BirdBones[BirdBones$type == "SO",]

```
```{r}
source("../scripts/BoneMeans.R")
dataMeans <- BoneMeans(BirdBones)
hist(SO$huml, breaks = 15)
```

```{r Swimming Birds}
boxplot(SW, main = "Bone Lenght Swimming Birds", 
        xlab = "Bone", ylab = "Length in mm", las=2)


```

    
    
    