---
# title: "Predicting Bird groups based on their bone measurements"
author: "Jouke Profijt"
date: "September 11, 2018"
output: pdf_document

---
```{r}
#Copyright (c) 2018 Jouke Profijt.
#Licensed under GPLv3. See gpl.md
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

```{r}
BirdBones <- read.csv("../data/bird.csv",header = T, sep = ",")
```

# Introduction
# Research Question
   What bone or group of bones that most birds have in common, is the most significant for the function in the diffrent ecological groups?
    
## Data
    Data recieved from:
    
[Birds' Bones and Living Habits, Kaggle dataset](https://www.kaggle.com/zhangjuefei/birds-bones-and-living-habits)
    
    Bone measurements were measured from a skeleton collection of Natural History Museum of Los Angeles County, 
    provided by Dr. D. Liu of beijing Museaum of Natural History
    
# Exploratory Data Analyses
The data contains 420 bird samples where the bone lengths and diameters have been measured. 
The birds are separated in 6 diffrent groups:

  * Swimming Birds, SW 
  * Wading Birds, W 
  * Terrestrial Birds, T 
  * Raptors, R 
  * Scansorial Birds, P 
  * Singing Birds, SO 

Most samples have data for:

  * Length and Diameter of the Humerus
  * Length and Diameter of the Ulna
  * Length and Diameter of the Femur
  * Length and Diameter of the Tibiotarsus
  * Length and Diameter of the Taesometatarsus


I'm creating a graph which displays the bonelengths on y axis an the Id on x colorcoded by their ecological group. by evaluating this we can see if some groups have overall larger or smaller bones and we see if there are big outliers.

```{r}
library(ggplot2)
library(reshape)
source("../scripts/BoneMeans.R")
BirdBones <- BoneMeans(BirdBones)
ggplot(data = BirdBones, aes(id, length.mean, colour = type)) +
  ggtitle("Bone lenghts per Ecological group")+
  ylab("Mean lenght of total bird bones in mm") +
  xlab("Bird Sample")+
  geom_point()
```
As seen above swimming birds have the biggest bones, but also shown is that there are a lot more samples in that group where there is a lot of variation. I can look into cleaning up the data and removing the biggest outliers in this group. Singing birds also have a lot of samples but there is much less variation and so more certanty. 

For the rest of the birds there are not a lot of sample so maby we could try and normalizing the data so there is an even amount of samples per group.

There are also 7 samples that contain missing values, we could just straight out not use these samples becouse 4 of these are part of the biggest group of samples. and the others are not part of the smallest groups.

```{r}
library(ggplot2)

ggplot(BirdBones, aes(x = type, y = length.mean, color = type)) +
  geom_boxplot()+
  coord_flip()+
  ggtitle("Boxplot for each ecological group's mean bone lenght")+
  ylab("Mean bone lenght in mm")+
  xlab("ecological group")
```

As you can see there are quite a few outliers in all groups except in group R ()

