---
title: "Quality Report"
author: "Jouke Profijt"
date: "September 30, 2018"
output: pdf_document
---
```{r}
#Copyright (c) 2018 Jouke Profijt.
#Licensed under GPLv3. See LICENSE
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pander)

```
```{r include=FALSE}
BirdBones <- read.csv("../data/bird.csv",header = T, sep = ",")
```

### TLDR;Summary
In my data cleaning i will be removing all samples with NA's becouse of the importance of these missing values. If the NA's were only in the smaller unimportant bones i would havee considered not removing them. I also took a look at the outliers which are the bigger birds, this data could be important but they are very diffrent from their respective means. I'm also not going to use bones with very little variation between them for classification, so the small bones, becouse: small variation makes for hard classification of bird groups.

```{r echo=FALSE}
NAs <- BirdBones[!complete.cases(BirdBones),]
huml.3rd.q <- 90.31
huml.1st.q <-25.17

out <- huml.1st.q - 1.5*(huml.3rd.q - huml.1st.q)
out.large <- huml.3rd.q + 1.5*(huml.3rd.q - huml.1st.q)
outliers <- subset(BirdBones, huml > out.large | huml < out)
summar <- as.table(c(nrow(NAs), nrow(outliers)))
row.names(summar) <- c("NA's", "Outliers")
pander(summar)
```
```{r}
# Below is some summary data to genarate the above conclusion
```
------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------


\newpage

```{r echo=FALSE}

pander(summary(BirdBones))
```

When we look at the summary we can see that every bone contains atleast 1 NA, And from our EDA we concluded that we dont want to use the small bones Femur & Tibiotarsus for classifiacation. So if the a sample only contains NA's in those two bones we can still use those samples. \newpage



```{r echo=FALSE}
# Table showing id's who are not complete.
pander(BirdBones[!complete.cases(BirdBones),])
```

we can see that there are no birds who only have NA's in the smaller bones so i suggest just not using the birds with NA's as shown above.

```{r Outliers, echo=FALSE}

pander(outliers)
```

As we can see most outliers come from the type Swimming Birds this could just mean that these birds are quite large but dont have a lot of samples to back it up.

```{r echo=FALSE}
a <- as.table(
c(
nrow(subset(BirdBones, type == "SW")),
nrow(subset(BirdBones, type == "W"))))

row.names(a) <- c("Swimming Birds", "Wading Birds")
pander(a, caption = "Sample count outliers")
```
```{r echo=FALSE}
means <- as.table(c(mean(subset(BirdBones, type == "SW")$huml),
                  mean(subset(BirdBones, type == "W")$huml)))





row.names(means) <- c("Swimming Birds", "Wading Birds")
pander(means, caption = "Mean lenghts")


```

As we can see here, all outliers are quite far from their respective means so that makes me consider that we should remove all outliers as they are in the groups with a lot of samples. 

