---
title: "Quality Report"
author: "Jouke Profijt"
date: "September 30, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pander)

```


```{r}
BirdBones <- read.csv("../data/bird.csv",header = T, sep = ",")
pander(summary(BirdBones))
```

When we look at the summary we can see that every bone contains atleast 1 NA, And from our EDA we concluded that we dont want to use the small bones Femur & Tibiotarsus for classifiacation. So if the a sample only contains NA's in those two bones we can still use those samples. \newpage



```{r}
# Table showing id's who are not complete.
pander(BirdBones[!complete.cases(BirdBones),])
```

we can see that there are no birds who only have NA's in the smaller bones so i suggest just not using the birds with NA's as shown above.

```{r Outliers}
huml.3rd.q <- 90.31
huml.1st.q <-25.17

out <- huml.1st.q - 1.5*(huml.3rd.q - huml.1st.q)
out.large <- huml.3rd.q + 1.5*(huml.3rd.q - huml.1st.q)

pander(subset(BirdBones, huml > out.large | huml < out))
```

As we can see most outliers come from the type Swimming Birds this could just mean that these birds are quite large but dont have a lot of samples to back it up.

```{r}
a <- as.table(
c(
nrow(subset(BirdBones, type == "SW")),
nrow(subset(BirdBones, type == "W"))))

row.names(a) <- c("Swimming Birds", "Wading Birds")
pander(a, caption = "Sample count outliers")
```
```{r}
means <- as.table(c(mean(subset(BirdBones, type == "SW")$huml),
                  mean(subset(BirdBones, type == "W")$huml)))





row.names(means) <- c("Swimming Birds", "Wading Birds")
pander(means, caption = "Mean lenghts")


```


Now we see that the wading birds outlier is a lot diffrent from its mean, while all swimming bird outliers are decently close and we must consider keeping those for clasification so we will remove the bird with id == 118, and we will remove all incomplete samples.

