---
title: "Predicting bird classifcation by bone lenghts"
author: "Jouke Profijt"
date: "October 8, 2018"
output:
  pdf_document: default
  html_document:
    df_print: paged
license: GPLv3
---
```{r echo=FALSE}
#Copyright (c) 2018 Jouke Profijt.
#Licensed under GPLv3. See LICENSE
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
# Classification of differing ecological bird populations

## Introduction

## Materials & Methods

## Resutls


### EDA
```{r echo=FALSE}
BirdBones <- read.csv("../data/bird.csv",header = T, sep = ",")
BirdBones.noNA <- BirdBones[complete.cases(BirdBones),]
length <- c(2,4,6,8,10)
library(ggplot2)

ggplot(stack(BirdBones.noNA[length]), aes(x = ind, y = values, color = ind)) +
  geom_boxplot()+
  xlab("Bone Type")+
  ylab("Length in mm")+
  labs(title = "Boxplot for bone lenghts per bone in mm", 
       subtitle = "Variation between small and large bones.",
       caption = "Graph 1: Bone lengths")
```
Graph 1 shows the average lenght per bone for all ecological groups. As we can see there are 2 bone types which are quite a bit smaller than the other 3 bones. The Femuer(green) is the smallest of these 2, this stands out because in us humans this is our biggest bone. after the Femeur the next smallest bone is the Taesometatarsus, this is expected as this bone connects the feet of the birds to their legs.

These 2 bones are quite small and show less variation than the other 3 bones. As we are trying to classify these bones they might be less important in our final classification algorithm. 

```{r, echo=FALSE}
source("../scripts/BoneMeans.R")
length.long <- c(2, 4, 8)
diameter.long <- c(3, 5, 9)
BirdBones.noNA.long <- BoneMeans(BirdBones.noNA, length.long, diameter.long)

ggplot(BirdBones.noNA.long,aes(x=log2(length.mean),y=log2(diameter.mean),color=type))+
  geom_point()+
  geom_smooth(method = loess)+
  xlab("Log2 Mean Lenght")+
  ylab("Log2 Mean Diameter")+
  labs(title = "Log2 transformed mean lenght vs mean diameter", subtitle="For Humerus, Ulna and Tibiotarsus", caption = "graph 2: lenght vs diameter"
       )
```


graph 2 shows us that the total bone lenght for the Swimming Birds is quite a bit larger than the other birds. Also we can see that the Terrestrial birds have thicker bones than the rest of the birds but are stull quite small. the singing birds are overall very small ans should be easy to classify. The raptors and Wading birds are a bit average and could become diffcult to classify but the raptors are a bit thicker and the Wading birds are a bit smaller.


## Data Cleaning
```{r, echo=FALSE}
library(ggplot2)
library(grid)
library(gridExtra)

 p1 <- ggplot(BirdBones.noNA.long, aes(x = type, y = length.mean, color = type)) +
  geom_boxplot()+
  ggtitle("Boxplot for each ecological group's mean bone lenght")+
  ylab("Mean bone lenght in mm")+
  xlab("ecological group")+
  labs("Boxplot for each ecological group's mean bone lenght", subtitle = "For Humerus, Ulna and Tibiotarsus",caption = "graph 3: mean bone lenghts")+
   theme(legend.position="none")





huml.3rd.q <- 90.31
huml.1st.q <-25.17

out <- huml.1st.q - 1.5*(huml.3rd.q - huml.1st.q)
out.large <- huml.3rd.q + 1.5*(huml.3rd.q - huml.1st.q)
outliers <- subset(BirdBones.noNA.long, huml > out.large | huml < out)
Birdbones.Clean <- BirdBones.noNA.long[! BirdBones.noNA.long$id %in% outliers$id, ]



p2 <- ggplot(Birdbones.Clean, aes(x = type, y = length.mean, color = type)) +
  geom_boxplot()+
  labs(
    title = "Boxplot for each ecological group's mean bone lenght",
    x = "ecological group",
    y = "",
    subtitle = "For Humerus, Ulna and Tibiotarsus",
    caption = "graph 4: Cleaned data"
  )+
  theme(legend.position="none")



grid.arrange(p1, p2, ncol = 2)
```
For our cleaning of data we removed all rows with NA's which cost us 7 datapoints. After that we calculated the 1st and 3rd quartile outliers from the humurus which can be seen in graph 3 & 4, and removed 9 rows. 

After all cleaning we left with 404 datapoints to use for Weka Analyses and classification.


## Weka
In weka the goal was to find a classification algorithem that has the highest possible accuracy because classifying a birds herritage it does not matter that much if there is a False positive, the goal was to keep false negatives low and keep true positives high.


```{r, include=FALSE}
library(pander)
 conf.matrix <-
   rbind(
    c(0,0,0,0,0,108),
    c(0,0,0,0,0,63),
    c(0,0,0,0,0,23),
    c(0,0,0,0,0,48),
    c(0,0,0,0,0,38),
    c(0,0,0,0,0,124))
colnames(conf.matrix) <- c("SW", "W", "T", "R", "P", "SO")

row.names(conf.matrix) <- c("SW", "W", "T", "R", "P", "SO")

pander(conf.matrix, caption = "Zero R Confusion matrix")

```

```{r, include=FALSE}
accuracy <- rbind(c(124, 280), c("30.6931 %", "69.3069 %"))

colnames(accuracy) <- c("Correct", "Incorrect")
rownames(accuracy) <- c("Instances", "Percentage")
pander(accuracy, caption = "Zero R classification results")
```


```{r, include=FALSE}
 conf.matrix <-
   rbind(
    c(69,11,1,10,8,9),
    c(35,2,4,4,10,8),
    c(4,1,4,2,8,4),
    c(29,5,1,6,6,1),
    c(5,0,7,0,16,10),
    c(2,0,3,0,10,109))
colnames(conf.matrix) <- c("SW", "W", "T", "R", "P", "SO")

row.names(conf.matrix) <- c("SW", "W", "T", "R", "P", "SO")

pander(conf.matrix, caption = "One R Confusion matrix")
```
```{r, include=FALSE}
accuracy <- rbind(c(208, 196), c("51.4851 %", "48.5149 %"))

colnames(accuracy) <- c("Correct", "Incorrect")
rownames(accuracy) <- c("Instances", "Percentage")
pander(accuracy, caption = "One R classification results")
```


### Chosen classifier: Random.Forest

```{r, echo=FALSE}
 conf.matrix <-
   rbind(
    c(87,11,0,4,0,6),
    c(17,34,0,2,4,6),
    c(2,0,12,2,4,3),
    c(7,1,0,37,3,0),
    c(0,0,1,1,30,6),
    c(0,0,1,1,1,121))
colnames(conf.matrix) <- c("SW", "W", "T", "R", "P", "SO")

row.names(conf.matrix) <- c("SW", "W", "T", "R", "P", "SO")

pander(conf.matrix, caption = "Random.Forest Confusion matrix as chosen classifier")
```


```{r, echo=FALSE}
accuracy <- rbind(c(321, 83), c("79.4554 %", "20.5446 %"))

colnames(accuracy) <- c("Correct", "Incorrect")
rownames(accuracy) <- c("Instances", "Percentage")
pander(accuracy, caption = "Random.Forest as chosen classifier")
```


```{r}
humurus <- signif(84.6535 - 82.1782, 4)
ulna <-  signif(84.6535 -83.1683, 4) 
Femur <-  signif(84.6535 -80.4455, 4)
Tibiotarsus <-  signif(84.6535 - 84.4059, 4) 
Taesometatarsus <-  signif(84.6535-83.1683, 4)

loss <- data.frame(Bone=c("humurus", "ulna", "Femur", "Tibiotarsus", "Taesometatarsus"), 
                   Classify_loss=c(humurus, ulna, Femur, Tibiotarsus, Taesometatarsus))

library(ggplot2)
ggplot(data=loss, aes(x=Bone, y=Classify_loss)) +
  geom_bar(stat="identity", fill = "blue")+
  geom_text(aes(label=Classify_loss), vjust=1.6, color="white", size=3.5)+
  theme_minimal()+
  labs(x = "bone type",
       y = "accuracy loss (%)",
       title = "Classification loss if certain bones are removed",
       subtitle = "For Random.forest",
       caption = "graph 5: accuracy loss per bone"
       )
```
in graph 5 we can see what the importance is from certain bones if we are talking about classification. the diffrences we can see are becouse of their importance for the functionallity of the bird groups.

```{r}
RemovedPercentage <- c(10, 20, 30, 40, 50, 60, 70, 80, 90)
Random.Forest <- c(22.07, 23.27, 25.17, 26.82, 29.17, 32.59, 35.48, 39.24, 47.05)
IBk <- c(18.26, 19.08, 20.63, 23.01, 25.26, 28.52, 31.58, 35.75, 45.99)

dataf <- data.frame(rev(RemovedPercentage), Random.Forest, IBk)
dataf <- dataf[seq(dim(dataf)[1],1),]

library(ggplot2)
ggplot(dataf)+
  geom_smooth(aes(x = RemovedPercentage, y = Random.Forest, color="Random.Forest"))+
  geom_smooth(aes(x = RemovedPercentage, y = IBk, color="IBk"))+
  labs(x = "Percentage of data",
       y = "False Positive rate",
       title = "Learning curve",
       subtitle = "For Random.Forest && IBk",
       caption = "graph 6: Learning curve inspected algorithm's"
       )
  
```

```{r}
rocdataW <- read.csv("../data/RocWading.arff", header = F, comment.char = "@", na.strings = "?")
rocdataWibk <- read.csv("../data/RocWadingIbk.arff", header = F, comment.char = "@", na.strings = "?")

library(ggplot2)
ggplot()+
  geom_line(aes(x = rocdataW$V6, y = rocdataW$V7, color="Random.Forest"))+
  geom_line(aes(x = rocdataWibk$V6, y = rocdataWibk$V7, color="IBk"))+
  labs(x = "Specificity",
       y = "Sensitivity",
       title = "ROC curve",
       subtitle = "Based on algorithm preformance on Wading bird classification"
       )
```

## Conclusion & discussion

What is the most important bone for each ecological group their function? we can conclude that (At least for classification) the Femur is the most important as when we remove this from our classification algorithems the accuracy loss is great. we can see this in graph 5 where the loss per bone is displayed. 
while researching the subject we made the assumption that the Femeur and Taesometatarsus were the least important for classification. Becouse of this the first classification algorithems were done only using the longer bones but further in reverted this desicion. The classification algorithem that was chosen however wasn't changed. In the future we should first collect some data about the importance for some bones before making such rational desisions.

We do wan to think about removing some bones as when we have less data to be put in the more unknown fossils with missing bones could possibly be classified

## Minor Proposal

## References
